Windowed least-squares model reduction for dynamical systems

### 2019-10-18 section4/spatial_numerical.tex
bug fixes

Direct approaches solve optimization problem~\eqref{eq:tclsrm} by
``transcribing" the infinite dimensional optimization problem into a finite
dimensional one by discretizing the state and objective functional in time.
The minimization problem is then reformulated as a (non)linear programming, or
optimization, problem. A variety of direct solution approaches exist, 
including collocation approaches, spectral  methods, and genetic algorithms.  
In the context of \spatialAcronym\ trial subspaces, the most straightforward direct solution approach consists of the 
following steps: (1) numerically discretize the FOM ODE (and hence the
\textit{integrand} of the objective function in problem~\eqref{eq:obj_gen_slab}) and 
(2) select a numerical quadrature rule to evaluate the \textit{integral} in~\eqref{eq:obj_gen_slab}.
To this end, we define time grids
$\{\timeWindowArg{n}{i}\}_{i=0}^{\nstepsArg{n}}\subseteq[\timeStartArg{n},\timeEndArg{n}]$ \EP{Shouldn't this be subset? We can't recover the continuous time interval},
$n=1,\ldots,\nslabs$ that
satisfy 
$\timeStartArg{n}=\timeWindowArg{n}{0} \le \cdots \le \timeWindowArg{n}{\nstepsArg{n}} 
 = \timeEndArg{n}$. %\KTC{use cdots, not
%hdots or ldots. Don't use $\defeq$ for the first and last times; these are
%just equalities, i.e., you happen to set them equal to those values, but they
%aren't identically the same:}
Figure~\ref{fig:slab_fig2} depicts such a discretization.
For the purposes of indexing between different windows, we additionally define a function:
$$\indexMapper: (n,i) \mapsto 
\begin{cases}
1,0 & n=1, \; i = 0,\\
n,i & n \ge 1, \; i > 0, \\
\indexMapper(n-1,\nstepsArg{n-1}+i) & n > 1, \; i \le 0.
\end{cases}$$
We now outline the direct solution approach for linear-multistep schemes; we note that the formulation for
other time-integration methods (e.g., Runge--Kutta) follows closely. 
\begin{figure} 
\begin{centering} 
\includegraphics[trim={0.0cm 4.5cm 0cm 3cm},clip,width=1.0\textwidth]{figs/time_grid_timesteps.pdf} 
	\caption{Depiction of the $\nstepsArg{n}+1$ time instances over each window. In the figure, $\nstepsArg{n} = 2$ for all $n$.} 
\label{fig:slab_fig2} 
\end{centering} 
\end{figure}
%We now outline several commonly employed temporal discretization 
%schemes, specifically linear multistep and collocation (Runge--Kutta) methods, that leverage this time-grid to discretize the FOM ODE and 
%objective functional.

@@

Employing a linear multistep method to discretize the FOM ODE yields the FOM O$\Delta$E over the $n$th window,

@@

&\residLMSWArg{n}{i} (\stateFOMDiscreteArg{n,i};\stateFOMDiscreteLMSArg{n}{i-1},\ldots,\stateFOMDiscreteLMSArg{n}{i-\lmsWidthArg{n}{i}}) = \bz, \qquad i=1,\ldots,\nstepsArg{n}, 

@@

In the above, $\stateFOMDiscreteArg{n,i} (\approx

@@

\objectiveArgLMS{n}_{\text{D}} &\vcentcolon (\stateyDiscreteArgnt{1},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}}; \stateyDiscreteArgnt{0},\ldots,\stateyDiscreteArgnt{-\lmsWidthArg{n}{1}+1}) \mapsto 

@@

\methodAcronym\ with the direct approach and a linear multistep method sequentially computes the solutions
$\approxstateDiscreteArg{n,1},\ldots,\approxstateDiscreteArg{n,\nstepsArg{n}}$,
$n=1,\ldots,\nslabs$, that satisfy
%	\KTC{What is $\genstate^{n-1}(\timeEndArg{n-1} ) $ doing here? Two problems
%	(1) there is no notion of a generalized coordinate $\genstate$ yet, and (2)
%	this should be the discrete solution, not the exact solution evaluated at
%	the final time.} \EP{You are correct, I think there is an error now in the minimize term though. It was $\in \trialspace\otimes \RR{\lmsWidthArg{n}{i}+1}$, it should be $\trialspace\otimes \RR{\nstepsArg{n}+1}$}
\begin{align*}
	&\underset{(\stateyDiscreteArgnt{1},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}}) \in ( \trialspace + \stateInterceptArg{n}) \otimes \RR{\nstepsArg{n}}}{\text{minimize } }
\objectiveArgLMS{n}_{\text{D-S}} (\stateyDiscreteArgnt{1},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}}), 
%%& \text{subject to} \hspace{0.2 in}  \stateyDiscreteArg{0} =
%%\begin{cases} \basisspaceArg{n} \spatialICDiscrete + \stateInterceptArg{n}& n = 2,\ldots,\nslabs,\\
%\basisspaceArg{1} \genstateICOne  + \stateInterceptArg{1}& n=1. \end{cases} 
\end{align*}
where
\begin{equation*}
\begin{split}
\objectiveArgLMS{n}_{\text{D-S}} &\vcentcolon  (\stateyDiscreteArgnt{1},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}}) \mapsto \objectiveArgLMS{n}(\stateyDiscreteArgnt{1},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}}; \projectorSpatialArg{n}(\approxstateDiscreteArg{\indexMapper(n,0)}),\ldots,\projectorSpatialArg{n}(\approxstateDiscreteArg{\indexMapper(n,-\lmsWidthArg{n}{1}+1)})), \\
&\vcentcolon \RR{\fomdim} \otimes \RR{\nstepsArg{n}} \rightarrow \RR{}_+,
\end{split}
\end{equation*}
and $\projectorSpatialArg{n} : \RR{\fomdim} \rightarrow \trialspaceArg{n}$ is the $\elltwo$-orthogonal projector 
$$\projectorSpatialArg{n} : \stateyDiscreteArgnt{} \mapsto  \basisspaceArg{n}\basisspaceTArg{n} (\stateyDiscreteArgnt{} - \stateInterceptArg{n}) + \stateInterceptArg{n} .$$
We additionally set $\approxstateDiscreteArg{1,0} = \projectorSpatialArg{n}(\stateFOMIC).
$
%where 
%\begin{align*}
%\residArg{n,i} &: (\stateyArgnt{n,i},\ldots,\stateyArgnt{n,i-k(i)}) \mapsto  \sum_{j=0}^k \alpha_j \stateyArgnt{n,i-j} = \Delta t^{n,i} \sum_{j=0}^k \beta_j \velocity(\stateyArgnt{n,i-j}),
%\\
%               &: \RR{\fomdim} \otimes \RR{k(i)} \rightarrow \RR{\fomdim}, 
%\end{align*} 
%is the discrete residual resulting from the linear multistep scheme and 
%\methodAcronym\ solves the minimization problem over each time window, 
%\begin{align*}
%&\approxstateDiscreteArg{n,0},\ldots,\approxstateDiscreteArg{n,\nstepsArg{n}} = \underset{\stateyDiscreteArgnt{n,0},\ldots,\stateyDiscreteArgnt{n,\nstepsArg{n}} \in \trialspace}{\text{arg\,min } }
%\objectiveArgLMS{n} (\stateyDiscreteArgnt{n,0},\ldots,\stateyDiscreteArgnt{n,\nstepsArg{n}}) \\
%& \text{subject to} \hspace{0.2 in}  \approxstateDiscreteArg{n,0} =
%\begin{cases} \approxstateDiscreteArg{n-1,\nstepsArg{n-1}} & n = 2,\ldots,\nslabs,\\
%\approxstateIC & n=1. \end{cases} \end{align*}

@@

\begin{remark}
For the limiting case where $\nstepsArg{n} = 1$, $n=1,\ldots,\nslabs$ such that the window size is equivalent to the time step (i.e., $\DeltaSlabArg{n} = \timeWindowArg{n}{1} - \timeWindowArg{n}{0}$), uniform 
quadrature weights are used, a uniform trial space is employed (i.e., $\trialspaceArg{n} = \trialspace$ and $\stateInterceptArg{n} = \stateInterceptArg{}, n=1,\ldots,\nslabs$), the weighting matrices are taken to be
$\stweightingMatOneArg{n} = \lspgWeightingArg{}, n=1,\ldots,\nslabs$, and the time instances $\timeWindowArg{n}{1} = t^n$, $n=1,\ldots,\nslabs$, then 
$\approxstateDiscreteArg{n,1}  = \approxstateLSPG^n$, $n=1,\ldots,\nslabs$ and  
\methodAcronym\ with \spatialAcronym\ trial
	subspaces solved via the direct approach recovers the LSPG approach. 
\end{remark} 
\subsubsection{Solution to the least-squares problem through the Gauss--Newton method}
	\KTC{I recommend rewriting this section with several important things in
	mind. We can discuss.}
	\KTC{If the dynamics are linear, doesn't this just yield a linear
	least-squares problem that can be solved directly? We need to present this
	more clearly.}
Discretization through linear multistep methods (as well as other techniques) 
results in a least-squares problem. In particular, for the case where the full-order model~\eqref{eq:FOM} is nonlinear, the direct approach yields a nonlinear least-squares problem. 
A variety of algorithms exist for solving nonlinear least-squares problems, including trust region approaches, the Gauss–Newton method, and the Levenberg–Marquardt method. The numerical experiments presented in this work comprise nonlinear systems and are solved via the Gauss--Newton method, and as such we outline this approach here.

>>>>

Direct approaches solve optimization problem~\eqref{eq:tclsrm} by
``transcribing" the infinite-dimensional optimization problem into a finite-dimensional one by discretizing the state and objective functional in time.
The minimization problem is then reformulated as a (non)linear 
optimization problem. A variety of direct solution approaches exist, 
including collocation approaches, spectral  methods, and genetic algorithms.  
In the context of \spatialAcronym\ trial subspaces, the most straightforward direct solution approach consists of the 
following steps: (1) numerically discretize the FOM ODE (and hence the
\textit{integrand} of the objective function in problem~\eqref{eq:obj_gen_slab}) and 
(2) select a numerical quadrature rule to evaluate the \textit{integral}
defining the objective function in~\eqref{eq:obj_gen_slab}.
To this end, we define time grids
$\{\timeWindowArg{n}{i}\}_{i=0}^{\nstepsArg{n}}\subset[\timeStartArg{n},\timeEndArg{n}]$
\EP{Shouldn't this be subset? We can't recover the continuous time interval}
\KTC{Yes},
$n=1,\ldots,\nslabs$ that
satisfy 
$\timeStartArg{n}=\timeWindowArg{n}{0} \le \cdots \le \timeWindowArg{n}{\nstepsArg{n}} 
 = \timeEndArg{n}$. %\KTC{use cdots, not
%hdots or ldots. Don't use $\defeq$ for the first and last times; these are
%just equalities, i.e., you happen to set them equal to those values, but they
%aren't identically the same:}
Figure~\ref{fig:slab_fig2} depicts such a discretization.
For the purposes of indexing between different windows, we additionally define a function:
$$\indexMapper: (n,i) \mapsto 
\begin{cases}
	(n,i) & n \geq 1, \; i \geq 0, \\
\indexMapper(n-1,\nstepsArg{n-1}+i) & n > 1, \; i \le 0.
\end{cases}$$
\KTC{Note: I added parenthesis here because a function can't really map to a
pair of indices unless they are defined as a collection or tuple. Should be
fine.}
We now outline the direct solution approach for linear-multistep schemes; the formulation for
other time-integration methods (e.g., Runge--Kutta) follows closely. 
\begin{figure} 
\begin{centering} 
\includegraphics[trim={0.0cm 4.5cm 0cm 3cm},clip,width=1.0\textwidth]{figs/time_grid_timesteps.pdf} 
	\caption{Depiction of the $\nstepsArg{n}+1$ time instances over each window. In the figure, $\nstepsArg{n} = 2$ for all $n$.} 
\label{fig:slab_fig2} 
\end{centering} 
\end{figure}
%We now outline several commonly employed temporal discretization 
%schemes, specifically linear multistep and collocation (Runge--Kutta) methods, that leverage this time-grid to discretize the FOM ODE and 
%objective functional.

@@

Employing such a method to discretize the FOM ODE yields the sequence of FOM
O$\Delta$Es defined over the $n$th time window

@@

&\residLMSWArg{n}{i} (\stateFOMDiscreteArg{n,i};\stateFOMDiscreteLMSArg{n}{i-1},\ldots,\stateFOMDiscreteLMSArg{n}{i-\lmsWidthArg{n}{i}}) = \bz, \qquad i=1,\ldots,\nstepsArg{n}

@@

Here, $\stateFOMDiscreteArg{n,i} (\approx

@@

\objectiveArgLMS{n}_{\text{D}} &\vcentcolon
	(\stateyDiscreteArgnt{1},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}};
	\stateyDiscreteArgnt{0},\ldots,\stateyDiscreteArgnt{-\lmsWidthArg{n}{1}+1})
	\mapsto 

@@

\methodAcronym\ with the direct approach and a linear multistep method sequentially computes the solutions
$\approxstateDiscreteArg{n,1},\ldots,\approxstateDiscreteArg{n,\nstepsArg{n}}$,
$n=1,\ldots,\nslabs$, that satisfy
%	\KTC{What is $\genstate^{n-1}(\timeEndArg{n-1} ) $ doing here? Two problems
%	(1) there is no notion of a generalized coordinate $\genstate$ yet, and (2)
%	this should be the discrete solution, not the exact solution evaluated at
%	the final time.} \EP{You are correct, I think there is an error now in the minimize term though. It was $\in \trialspace\otimes \RR{\lmsWidthArg{n}{i}+1}$, it should be $\trialspace\otimes \RR{\nstepsArg{n}+1}$}
\begin{align*}
	&\underset{(\stateyDiscreteArgnt{1},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}}) \in ( \trialspace + \stateInterceptArg{n}) \otimes \RR{\nstepsArg{n}}}{\text{minimize } }
\objectiveArgLMS{n}_{\text{D-S}} (\stateyDiscreteArgnt{1},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}}), 
%%& \text{subject to} \hspace{0.2 in}  \stateyDiscreteArg{0} =
%%\begin{cases} \basisspaceArg{n} \spatialICDiscrete + \stateInterceptArg{n}& n = 2,\ldots,\nslabs,\\
%\basisspaceArg{1} \genstateICOne  + \stateInterceptArg{1}& n=1. \end{cases} 
\end{align*}
where
\KTC{This next equation: I don't think you should project the previous
solutions.}
\begin{equation*}
\begin{split}
\objectiveArgLMS{n}_{\text{D-S}} &\vcentcolon  (\stateyDiscreteArgnt{1},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}}) \mapsto \objectiveArgLMS{n}(\stateyDiscreteArgnt{1},\ldots,\stateyDiscreteArgnt{\nstepsArg{n}}; \projectorSpatialArg{n}(\approxstateDiscreteArg{\indexMapper(n,0)}),\ldots,\projectorSpatialArg{n}(\approxstateDiscreteArg{\indexMapper(n,-\lmsWidthArg{n}{1}+1)})), \\
&\vcentcolon \RR{\fomdim} \otimes \RR{\nstepsArg{n}} \rightarrow \RR{}_+,
\end{split}
\end{equation*}
and $\projectorSpatialArg{n} : \RR{\fomdim} \rightarrow \trialspaceArg{n}$ is the $\elltwo$-orthogonal projector 
$$\projectorSpatialArg{n} : \stateyDiscreteArgnt{} \mapsto  \basisspaceArg{n}\basisspaceTArg{n} (\stateyDiscreteArgnt{} - \stateInterceptArg{n}) + \stateInterceptArg{n} .$$
We additionally set $\approxstateDiscreteArg{1,0} = \projectorSpatialArg{n}(\stateFOMIC).
$
%where 
%\begin{align*}
%\residArg{n,i} &: (\stateyArgnt{n,i},\ldots,\stateyArgnt{n,i-k(i)}) \mapsto  \sum_{j=0}^k \alpha_j \stateyArgnt{n,i-j} = \Delta t^{n,i} \sum_{j=0}^k \beta_j \velocity(\stateyArgnt{n,i-j}),
%\\
%               &: \RR{\fomdim} \otimes \RR{k(i)} \rightarrow \RR{\fomdim}, 
%\end{align*} 
%is the discrete residual resulting from the linear multistep scheme and 
%\methodAcronym\ solves the minimization problem over each time window, 
%\begin{align*}
%&\approxstateDiscreteArg{n,0},\ldots,\approxstateDiscreteArg{n,\nstepsArg{n}} = \underset{\stateyDiscreteArgnt{n,0},\ldots,\stateyDiscreteArgnt{n,\nstepsArg{n}} \in \trialspace}{\text{arg\,min } }
%\objectiveArgLMS{n} (\stateyDiscreteArgnt{n,0},\ldots,\stateyDiscreteArgnt{n,\nstepsArg{n}}) \\
%& \text{subject to} \hspace{0.2 in}  \approxstateDiscreteArg{n,0} =
%\begin{cases} \approxstateDiscreteArg{n-1,\nstepsArg{n-1}} & n = 2,\ldots,\nslabs,\\
%\approxstateIC & n=1. \end{cases} \end{align*}

@@

\begin{remark}
For the limiting case where $\nstepsArg{n} = 1$, $n=1,\ldots,\nslabs$ such that the window size is equivalent to the time step (i.e., $\DeltaSlabArg{n} = \timeWindowArg{n}{1} - \timeWindowArg{n}{0}$), uniform 
quadrature weights are used, a uniform trial space is employed (i.e., $\trialspaceArg{n} = \trialspace$ and $\stateInterceptArg{n} = \stateInterceptArg{}, n=1,\ldots,\nslabs$), the weighting matrices are taken to be
$\stweightingMatOneArg{n} = \lspgWeightingArg{}, n=1,\ldots,\nslabs$, and the time instances $\timeWindowArg{n}{1} = t^n$, $n=1,\ldots,\nslabs$, then 
$\approxstateDiscreteArg{n,1}  = \approxstateLSPG^n$, $n=1,\ldots,\nslabs$ and  
\methodAcronym\ with \spatialAcronym\ trial
	subspaces solved via the direct approach recovers the LSPG approach. 
\end{remark} 
\subsubsection{Solution to the least-squares problem through the Gauss--Newton
	method}
Discretization through linear multistep methods (as well as other techniques) 
results in a least-squares problem. In particular, for the case where the full-order model~\eqref{eq:FOM} is nonlinear, the direct approach yields a nonlinear least-squares problem. 
A variety of algorithms exist for solving nonlinear least-squares problems, including trust region approaches, the Gauss–Newton method, and the Levenberg–Marquardt method. The numerical experiments presented in this work comprise nonlinear systems and are solved via the Gauss--Newton method, and as such we outline this approach here.
